% Chapter:  Introduction to point counting  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Varieties over finite fields and zeta functions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Varieties over finite fields and zeta functions}

Let $X$ denote an algebraic variety over the finite field~$\mathbf{F}_q$ 
with $q = p^a$ elements.  In particular, $X$ is also defined 
over every algebraic extension $\mathbf{F}_{q^i}$, for $i \geq 0$, 
and we can let $N_i = \abs{ X(\mathbf{F}_{q^i}) }$ denote the number 
of points of $X$ over $\mathbf{F}_{q^i}$.  The \emph{zeta function} 
of $X$ is then defined as the formal power series in $\mathbf{Q}[[T]]$,
\begin{equation*}
Z(X, T) = \exp \Bigl( \sum_{i=0}^{\infty} \frac{N_i T^i}{i} \Bigr).
\end{equation*}
In the above form, the power series~$Z(X, T)$ is represented by an 
infinite series and it is unclear whether this can be explicitly 
computed by an algorithm viz.\ a \emph{finite} routine.  This issue 
was resolved by Dwork~\citep{Dwork1960} in 1960:

\begin{thm}
Let $X$ be an algebraic variety over the finite field~$\mathbf{F}_q$.  
Then the zeta function $Z(X, T)$ is a quotient of two polynomials with 
integer coefficients.
\end{thm}

Moreover, Bombieri~\citep{Bombieri1978} provided an explicit bound 
for the total degree of the rational function representing $Z(X, T)$.
Dwork's theorem is included in a group of results known as the 
\emph{Weil conjectures}, statements conjectured by Weil~\citep{Weil1949} 
in~1949 which have subsequently been proved.  We follow 
Kedlaya~\citep[Theorem~1.2.1]{Kedlaya2011} in the statement of the Weil 
conjectures:

\begin{thm} \label{thm:01-Zetafunctions}
Let $X$ be a separated scheme of finite type over the finite 
field~$\mathbf{F}_q$.  Then the zeta function of~$X$ is the power series 
representation of a rational functions~$T$.  Moreover, if $X$ is smooth 
and proper over~$\mathbf{F}_q$, then there is a unique way to write 
\begin{equation}
Z(X, T) = \prod_{i=0}^{2 \dim(X)} P_i(T)^{(-1)^{i+1}}
\end{equation}
for some polynomials $P_i(T) \in \mathbf{Z}[T]$ with $P_i(0) = 1$, 
satisfying the following conditions.
\begin{enumerate}
\item We have 
\begin{equation*}
P_i(q^{-i} / T) = \pm q^{-i \deg(P_i) / 2} T^{- \deg(P_i)} P_i(T),
\end{equation*}
with the sign being $+$ whenever $i$ is odd.  In other words, the roots 
of~$P_i$ are invariant under the map $r \mapsto q^{-i} / r$, and if $i$ 
is odd then the multiplicities of $\pm q^{-i/2}$ are even.
\item The complex roots of $P_i$ all have absolute value $q^{-i/2}$.
\item If $X \cong \mathfrak{X}_{\mathbf{F}_q}$ for some smooth proper 
scheme~$\mathfrak{X}$ over the local ring $R = \mathfrak{o}_{K,\mathfrak{p}}$, 
for some number field~$K$ and some prime ideal~$\mathfrak{p}$ of 
$\mathfrak{o}_K$ with residue field~$\mathbf{F}_q$, then for any 
embedding~$K \into \mathbf{C}$, 
\begin{equation*}
\deg(P_i) = \dim_{\mathbf{C}} H^{i} \bigl( (\mathfrak{X} \times_R \mathbf{C})^{\text{an}}, \mathbf{C} \bigr).
\end{equation*}
In other words, $\deg(P_i)$ equals the $i$-th Betti number of 
$\mathfrak{X} \times_{R} \mathbf{C}$.
\end{enumerate}
\end{thm}

A discussion of the proof of this theorem is significantly 
beyond the scope of this thesis.  As a starting point for 
further details, we refer the reader to 
Hartshorne~\citep[Appendix~C]{Har77} and Osserman~\citep{Osserman2008}.

Dwork's theorem on the rationality of the zeta function implies 
that $Z(X, T)$ is already defined by a finite initial segment of 
the sequence~$(N_i)_{i \geq 0}$.  In particular, the problem of 
computing the zeta function can be formulated as follows:

\begin{prob} \label{prob:dm-pointcounting}
Given an algebraic variety~$X$ defined by a finite list of 
multivariate polynomials in~$\mathbf{F}_{q}[x_0, \dotsc, x_n]$, 
efficiently compute a rational function in~$\mathbf{Q}(T)$ 
representing $Z(X, T)$.
\end{prob}

% Point counting algorithms %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Point counting algorithms}

Computational routines addressing Problem~\ref{prob:dm-pointcounting} are 
known as \emph{point counting} algorithms.  Perhaps the most naive approach 
is to explicitly exhibit the points of~$X$ over~$\mathbf{F}_{q^i}$ by an 
exhaustive search for sufficiently many $i \geq 0$, determining $Z(X, T)$.  
For example, in the case of an elliptic curve~$E$ over a finite 
field~$\mathbf{F}_{q}$, the zeta function is given by 
\begin{equation*}
Z(E, T) = \frac{1 - a_q T + q T^2}{(1 - T)(1 - q T)}
\end{equation*}
where $a_q = q + 1 - \abs{E(\mathbf{F}_{q})}$.  However, this 
approach is only feasible for the smallest of instances of 
Problem~\ref{prob:dm-pointcounting}.  More sophisticated algorithms 
have been developed, and we refer the reader to a survey by 
Chambert-Loir~\citep{ChambertLoir2008} on point counting algorithms 
in general  and a survey by Kedlaya~\citep{Kedlaya2004} on $p$-adic 
methods in particular.  In the following, we only mention a few 
different approaches.

In~1985, Schoof~\citep{Schoof1985} proposed an $\ell$-adic algorithm for 
computing the zeta function of an elliptic curve~$E$.  The main idea behind 
this approach is to compute the trace~$a_q \bmod{\ell}$ of the Frobenius 
endomorphism on $H_{\text{\'et}}^{1}(E, \mathbf{F}_{\ell})$ for sufficiently 
many primes~$\ell$ and then determine the integer~$a_q$ using the Chinese 
remainder theorem.  Schoof's algorithm has subsequently been improved by 
Atkin~\citep{Atkin1988,Atkin1992} and Elkies~\citep{Elkies1991,Elkies1998}, 
leading to a practical algorithm with runtime polynomial in $\log q$.  
It has also been generalised, for example to higher genus curves by 
Pila~\citep{Pila1990}, however, the runtime remains exponential in the genus.

Another group of algorithms applicable primarily for elliptic curves 
is based on the \emph{canonical lift} of a curve from~$\mathbf{F}_{q}$ 
to the ring~$\mathbf{Z}_{q}$, which is the ring of integers of the 
unique unramified extension of $\mathbf{Q}_{p}$ with residue field 
$\mathbf{F}_{q}$.  The first of these algorithms is Satoh's 
method~\citep{Satoh2000} presented in~2000, which has runtime 
polynomial in the extension degree~$\log_{p} q$ but linear in the prime~$p$.

Finally, there is a class of so called $p$-adic algorithms based on 
Kedlaya's algorithm~\citep{Kedlaya2001}, which directly computes a 
$p$-adic approximation to the action of Frobenius on the rigid cohomology 
spaces associated to hyperelliptic curves in odd characteristic.  
For a fixed prime~$p$, the runtime of Kedlaya's original algorithm 
is polynomial in the extension degree~$\log_{p} q$ and the genus~$g$.  
However, there have been a number of improvements.  Harvey~\citep{Harvey2007} 
improved the roughly linear dependence on the prime from~$p$ to $\sqrt{p}$.  
Denef and Vercauteren~\citep{DenVer2006} included the case $p = 2$, and 
Castryck, Denef and Vercauteren~\citep{CasDenVer2006} extended the algorithm 
to so called nondegenerate curves.

Lauder's \emph{deformation method} presented in~\citep{Lau04a} 
applies to smooth projective hypersurfaces, additionally making use 
of the theory of deformation due to Dwork~\citep{Dwork62b}.  The 
runtime of this algorithm is polynomial in the prime~$p$, the 
extension degree~$\log_{p} q$, and the genus~$g$.  Subsequently, 
Gerkmann~\citep{Gerkmann2007} improved a number of $p$-adic bounds 
and developed a better implementation.  Moreover, the practicality 
of the deformation method has also been demonstrated for various 
classes of curves.  Hubrechts~\citep{Hubrechts2007, Hubrechts2008} 
considered hyperelliptic curves, Castryck, Hubrechts and 
Vercauteren~\citep{CasHubVer2008} considered a slightly 
more general class of $C_{a,b}$-curves, and Tuitman~\citep{Tuitman2011} 
considered nondegenerate curves.  These results provide the context 
for this thesis, in which we demonstrate the practicality for surfaces 
and higher-dimensional hypersurfaces.

Finally, we mention Lauder's \emph{fibration method} introduced 
in~\citep{Lauder2006}, which proceeds recursively on the dimension 
of a hypersurface.  This approach has been developed further 
by Walker~\citep{Walker2009}.

